---
title: "Writing your own functions in R"
author: "Will Cornwell, Daniel Falster"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

# Introduction

The goal of this session is to learn how to write your own functions in R. 

Three big reasons to write functions

1. Using function to reuse code / avoid repeated code
2. Using function to tidy up your analysis
3. Using (anonymous) functions to extend possibilities in dplyr or map workflows 

Research is by definition new, and so sometimes you have to do things that no one has done before.  If you do that new thing more than once, it's worth writing it into a function

Use similar packages and data to yesterday 
```{r}
library(tidyverse)
library(performance)
```

## First, let's load the dataset

```{r}
data <- list.files("data/penguins/", full.names = TRUE) |> map_df(read_csv)
```


## Now some examples on what we can achieve

Example 1 - tidy & turbo charge model fitting

```{r}
only_counts <- filter(data, individualCount > 0)

fit_model <- function(data_in) {
  lm(log10(individualCount) ~ decimalLatitude, data = data_in)
}



get_r2 <- function(fit_in) {
  r <- performance::r2(fit_in)
  return(r$R2)
}

fits <-
  only_counts |>
  split(~species) |>
  map(fit_model)

performance::check_model(fits[[4]])

map(fits, get_r2)
```


Example 2 - tidy & turbo charge plotting

```{r}
make_plot <- function(penguin_of_interest, pen_all = pen_all) {
  world <- map_data("world")
  specific_pen <- filter(pen_all, species == penguin_of_interest)
  pen_large_counts <- filter(specific_pen, individualCount >= 1000)
  pen_small_counts <- filter(specific_pen, individualCount < 1000)
  worldplot <- ggplot() +
    geom_polygon(data = world, aes(x = long, y = lat, group = group)) +
    coord_fixed(1.3) +
    geom_point(
      data = pen_small_counts,
      aes(x = decimalLongitude, y = decimalLatitude),
      col = "grey",
      alpha = 0.1
    ) +
    geom_point(
      data = pen_large_counts,
      aes(x = decimalLongitude, y = decimalLatitude),
      col = "red"
    ) +
    ggtitle(penguin_of_interest) +
    theme_void()
  print(worldplot)
  return(worldplot)
}


make_plot(unique(data$species)[10], data)
```


Now let's use our new skills to make lots of plots!

```{r}
penguin_species <- unique(data$species)
walk(penguin_species, make_plot, data)
```

Or we could save the plots
```{r}
save_plot <- function(penguin_of_interest_in, pen_all, path = "output/by_species/") {
  dir.create(path, FALSE, TRUE)
  make_plot(penguin_of_interest_in, pen_all)
  ggsave(paste0(path, penguin_of_interest, ".png"))
}

for (penguin_of_interest in penguin_species) {
  save_plot(penguin_of_interest, data)
}
```

# Structure of a function

function structure (rowSums)

- name
- body
- arguments

investigating structure of function f 

- `args(f)`
- `body(f)` or just `f`
- return value
- ?help
- named vs ordered arguments
- default arguments

Here's a simple example, implementing a simple function

```{r}
mymean <- function(x) {
  sum(x) / length(x)
}
```

**Exercises:**  

Load penguin data
```{r}
data <- list.files("data/penguins/", full.names = TRUE) |> map_df(read_csv)
```

1. Extract the most northern observations for each species in the southern hemisphere Write a function that finds the largest value that is below 0

```{r, drop}
find_largest_negative <- function(x) {
  max(x[x < 0])
}

data |>
  group_by(scientificName) |>
  summarise(most_northern_south = find_largest_negative(decimalLatitude))
```

2. If your function returns -Inf for some cases, change what it returns to NA possibly using `na_if` or `ifelse` inside the function.  This is called handling edge cases ellegantly.  Of course this can be done outside the function, but it's a better, more robust function if this is done internally.  

```{r, drop}
find_largest_negative_better <- function(x) {
  y <- max(x[x < 0])
  na_if(y, -Inf)
}

# OR

find_largest_negative_better <- function(x) {
  if (all(x > 0)) {
    y <- NA
  } else {
    y <- max(x[x < 0], na.rm = TRUE)
  }
  return(y)
}


data |>
  group_by(species) |>
  summarise(
    north_south = find_largest_negative_better(decimalLatitude)
  )
```


# Combining functions with map

From yesterday, we had this example

```{r}
list.files("data/species", full.names = TRUE) |> map_df(read_csv)
```

What if we want to record which files each data point came from? Write a function that loads a file and adds the filename in a new column of the data

```{r, drop}
load_file <- function(x) {
  read_csv(x) |> mutate(filename = x)
}

data <- list.files("data/penguins/", full.names = TRUE) |> map_df(read_csv)
```


**Exercises:**  We'll now make some plots

some pretty code to use 

```{r}
world_map <- map_data("world")
ggplot() +
  geom_polygon(
    data = world_map, aes(x = long, y = lat, group = group),
    fill = "darkgrey", colour = "white"
  ) +
  coord_equal()
```

1. Write a function that takes the dataframe and a genus name and plots the distribution for *a given genus* and saves it to the output file in a folder called "figures". The first line will create a folder.  

```{r}
dir.create("figures")
```

Try here:

```{r, drop}


plot_genus <- function(df, genus) {
  df_specific_family <-
    filter(df, genus == genus)
  world_map <- map_data("world")
  ggplot(df_specific_family, aes(x = decimalLongitude, y = decimalLatitude)) +
    geom_polygon(
      data = world_map, aes(x = long, y = lat, group = group),
      fill = "darkgrey", colour = "white"
    ) +
    geom_point() +
    ggtitle(genus)
  ggsave(paste0("figures/", genus, ".pdf"))
}

plot_genus(data, "Eudyptes")
```

2. Now re-write this function so it works with `map`, and make a plot of all the genera

```{r, drop}
plot_genus <- function(df, genus) {
  df_specific_family <-
    filter(df, genus == genus)
  world_map <- map_data("world")
  ggplot(df_specific_family, aes(x = decimalLongitude, y = decimalLatitude)) +
    geom_polygon(
      data = world_map, aes(x = long, y = lat, group = group),
      fill = "darkgrey", colour = "white"
    ) +
    geom_point() +
    ggtitle(genus)
  ggsave(paste0("figures/", genus, ".pdf"))
}

genus_list <- as.list(unique(data$genus))

map(genus_list, plot_genus, df = data)
```


3. Your penguin colleagues seem to think that some species migrate more than others but no one knows by how much.  They ask you to write a function that calculates the difference between median jun-jul-aug-sept latitude and median dec-jan-feb-mar latitude.  Then apply this function to each species. Return a table of "migration distance in latitude degrees" by species, also include the number of observations in summer and the number in winter.  the `%in%` function may be useful.  

Note some penguin species may have never been observed in winter.  (Where do they go????)

```{r drop}
migration_latitude <- function(species_of_interest, df) {
  sp_df <- filter(df, species == species_of_interest)
  winter <- filter(sp_df, month %in% c(6, 7, 8, 9))
  summer <- filter(sp_df, month %in% c(12, 1, 2, 3))
  tibble(
    species = species_of_interest,
    lat_migration = median(summer$decimalLatitude, na.rm = TRUE) - median(winter$decimalLatitude, na.rm = TRUE),
    winter_n = nrow(winter),
    summer_n = nrow(summer)
  )
}

output_df <- list()
for (i in seq_along(unique(data$species))) {
  output_df[[i]] <- migration_latitude(unique(data$species)[i], data)
}
bind_rows(output_df)
```

## Anonymous functions

Remember we did this:

```{r}
load_file <- function(x) {
  read_csv(x) |> mutate(filename = x)
}

list.files("data/species", full.names = TRUE) |> map_df(load_file)
```


Another way:

```{r}
list.files("data/species", full.names = TRUE) |>
  map_df(function(x) read_csv(x) |> mutate(filename = x))


# Or
list.files("data/species", full.names = TRUE) |>
  map_df(~ read_csv(.x) |> mutate(filename = .x))
```


**Exercises:** Reimplement the following example as an anonymous function

```{r, drop}
short_name <- function(x) {
  x2 <- x |> str_split(" ")
  paste(x2[[1]][1], x2[[1]][2])
}

data$scientificName |> map_chr(short_name)
```

## Extra exercises

Visualize penguin migrations for the different species.  one approach involves re-writing the function above to get out median lat and long.  

```{r}
```

